# Preprocess of Functional Analysis of KEGG

(This script was written in _**/share/projects/SOP/Functional_Analysis/github/Functional_analysis/**_ , all relative paths below are under this complete path.)

_**This demo guides you how to merge and generate KO profile and pathway profile from humann2 pipeline output folder in bash.**_  

In this Chapter, data of a cohort consisting of 10 samples from 8 patients, 1 water tube and 1 BRS in Newtech project were used as demo data.  

---

## Environment Setup

```{r Environment Setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

---

## Process pipeline output

_**Demo datacan be found in KEGG/Demo_data**_

- Pipeline_Out. Folder containing all samples' pipeline result.

- metadata.xls. Metadata of samples.


_**Folder structure of humann2 pipeline result**_  

Four files should be included in humann2_kegg folder of each sample:

1. *_genefamilies.tsv (KEGG gene profile file, rows are gene names consist of SpeciesCode:GeneName, columns are samples, unit of abundance is RPK)

2. *.log

3. *_pathabundance.tsv

4. *_pathcoverage.tsv

```{bash folder structure}
tree -L 4 ./Demodata/KEGG/Demo_data/Pipeline_Out/14286/
```

_**Merge data from humann2 pipeline (Kegg database)**_

Use Script merge_metaphlan_tables.py from humann2 to merge all gene_family profile files.

In this chunk, we merge all *genefamilies.tsv into one file and edit the colnames of the table

- input: *genefamilies.tsv, path of all *genefamilies.tsv files

- output: file name of merged profile table.

```markdown
## Start merging profile files

docker run -i --rm -u $(id -u):$(id -g) -v /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/:/in harbor.xbiome.com/xbiome/environments/humann2:2.8.1-2b8c5c3 bash -c "merge_metaphlan_tables.py /in/Demo_data/Pipeline_Out/*/humann2_kegg/*genefamilies.tsv > /in/Demo_data/merged_profile.tsv"

## Remove suffix in sample name
sed -i 's#_genefamilies##g' /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged_profile.tsv
sed -i 's$^ID$# Gene Family$g' /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged_profile.tsv

```

_**Transform all species code with species name in merged profile file & map kegg gene ID to KO entries**_

In this chunk, we transform the species code to species names and convert the kegg gene id to K numbers.

- input1: *profile.tsv  

- input2: path of kegg database

- output: *kegg, renamed profile table. (suffix must be *.kegg)

```markdown
## Mapping file prokaryotes.dat.gz was used to map species code to species names and map kegg gene id to K numbers.
## Version of prokaryotes.dat.gz should be consistent with the version of Kegg database.
## Note!!!! file name of output must ends with ".kegg"

python ./Scripts/rename-kegg-gene.py -in /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged_profile.tsv -db ./Scripts/db/ -out /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged.kegg -species 1

tail -n 2 /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged.kegg
```

_**Remove unmapped row in merged profile file**_

In this chunk, we remove unmapped row from merged profile file.

```markdown
## Since unmapped row doesn't convey any information we need, we remove unmapped row from merged profile file.

grep -v "^UNMAPPED" /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged.kegg > /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged_remove_unmapped.txt
```

_**Re-scale (Total Sum Scale) & map kegg gene entries to KO entries**_

In this chunk, we re-scale the profile of each feature by divding their RPK value by the column sum (sum of RPK values within one sample).

- input: the *remove_unmapped.txt file from last step

- output1: *knumber.txt, the profile table of all k numbers in all samples

- output2: *input, a list including all K numers in *knumber.txt

```markdown
## In this chunk we take merged_remove_unmapped.txt from last step as input, re-scale all features abundance into relative abundance (sum to 1). We also extracted the abundance of K numbers and saved in *knumber.txt

python ./Scripts/reminpath.py -i /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged_remove_unmapped.txt -k /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged_knumber.txt -m /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged.input
```

_**Use minpath to extract the mapping information between Knumers and KEGG pathways**_

In this chunk, we use minpath to extract the mapping relation between K numers and KEGG pathways and generate *report and *details for the calculation of pathway abundance.

- input: *input from last step

- output1: *report, report of minpath

- output2: *details, detail of minpath

```markdown
python /share/work/runtime/softwares/MinPath/MinPath.py -ko /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged.input -report /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged.report -details /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged.details > /dev/null 2>&1
```

_**Calculate the abundance of Kegg pathways**_

In this chunk we calculate the abundance of KEGG pathways. Since multiple K numbers may belong to the same Pathway, we provided three methods to calculate the abundance of pathway from  multiple K numbers:

1. Harmonic Mean

2. Median

3. Average of top 50% abundance KOs

- input1: *report from last step

- input2: *details from last step

- input3: *remove_unmapped.txt from chunk "remove unmapped row", unit for each feature is RPK.

- output: *pathway.txt, pathway profile file

```markdown
## In this chunk, we use the minpath report (stroing the mapping information between KO and Kegg pathway) and abundance table (unit of feature is RPK) of KO to estimate the abundance of pathways.
## Note: Becasue minpath mapping file doesn't include the UMAPPED entry, we need to add UMAPPED entry to the output manually after the calcualtion of pathway abundance.

python ./Scripts/kegg_abundance.py -path /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged.report -detail /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged.details -in  /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged_remove_unmapped.txt -out /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged_pathway.txt

## Add UNMAPPED row to pathway abundance table

grep "UNMAPPED" /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged.kegg | grep -v "|" >> /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged_pathway.txt

## Quick view of kegg pathway profile
head /share/projects/SOP/Functional_Analysis/github/Functional_analysis/Demodata/KEGG/Demo_data/merged_pathway.txt
```

---

## Session info  

```{r}
devtools::session_info()
```  